---
title: "Keeling Curve"
subtitle: ""
title-block-banner-color: "#6c757d"
title-block-banner: "#E5E5E5"
format: 
  html:
    toc: true
    toc-depth: 3
    toc-title: "Report contents"
    code-link: true
    highlight-style: a11y
    link-external-newwindow: true
    embed-resources: true
execute: 
  message: false
  warning: false
bibliography: references.bib
---

## Introduction

This report processes and visualises data using [R](https://www.r-project.org/) and [Observable JS](https://observablehq.com/@observablehq/observables-not-javascript) to demonstrate some of the functionality of [Quarto](https://quarto.org/).

## Data

The [data](https://scrippsco2.ucsd.edu/data/atmospheric_co2/primary_mlo_co2_record.html) derive from the Scripps Institution of Oceanography at UC San Diego [@keeling2005atmospheric].

## Load packages

::: panel-tabset
### R

```{r}
#| label: load-packages

library(readr)
library(dplyr)
library(lubridate)
```

### OJS

```{ojs}
//| label: load-libraries

Plot = import("https://cdn.jsdelivr.net/npm/@observablehq/plot/+esm")
d3 = require("d3@7")
```
:::

## Load data

```{r}
#| label: load-data

df <- read_csv("https://scrippsco2.ucsd.edu/assets/data/atmospheric/stations/in_situ_co2/monthly/monthly_in_situ_co2_mlo.csv", skip = 63, na = "-99.99") |> 
  select(date = 4, average = 5) |> 
  filter(!is.na(average)) |> 
  mutate(date = as.Date(date_decimal(date, tz = "UTC"))) |>
  arrange(date)

ojs_define(df = df)
```

## Inspect data

@tbl-readings shows the latest readings taken from the [Mauna Loa Observatory](https://gml.noaa.gov/obop/mlo/) in Hawaii.

```{r}
#| label: tbl-readings
#| tbl-cap: "Atmospheric CO₂ during 2024"
#| code-fold: true

library(gt)

df |> 
  arrange(desc(date)) |> 
  gt() |> 
   cols_label(
    date = "Date",
    average = "CO₂ (ppm)"
  ) |> 
  fmt_date(
    columns = date,
    date_style = "yMMMM"
  ) |> 
  tab_style(
    style = cell_text(align  = "right", weight = "bold"),
    locations = cells_column_labels()
  ) |> 
  tab_options(
    table.align = "left"
  ) |> 
   opt_interactive()
```

## Plot data

@fig-co2 is the [Keeling Curve](https://keelingcurve.ucsd.edu/), an iconic line chart showing changes in atmospheric levels of carbon dioxide since 1958. CO<sub>2</sub> levels are currently rising at around 2.5 parts per million (ppm) each year.

```{ojs}
//| label: transform-data

data = transpose(df).map(d => ({
  date: new Date(d.date),
  average: +d.average
}))
```

```{ojs}
//| code-fold: true
//| label: fig-co2
//| fig-alt: | 
//|   Line chart showing atmospheric concentrations of carbon dixide between 1958 and the present.  
//| fig-cap: |
//|   Monthly average CO₂ atmospheric concentrations

Plot.plot({
  height: 396, width: 640,
  marginRight: 50,
	x: {
    label: "",
    labelArrow: null
    },
  y: {
    domain: [300, d3.max(data, d => d.average)],
		ticks: 5,
    tickSize: 0,
    tickPadding: 3,
    axis: "right",
    grid: true,
    label: "CO₂ concentration (ppm)",
    labelArrow: null
    },
  marks: [
    Plot.lineY(data, {
      x: "date", 
      y: "average",
      stroke: "#636363"
      }),
    Plot.tip(data, Plot.pointer({
      x: "date",
      y: "average",
      title: (d) => [`${d3.timeFormat("%b %Y")(d.date)}\n${(d.average)}ppm`].join("   ")
    })),
    Plot.ruleY([300])
    ],
    style: {fontSize: "14px", fontFamily: "'Open Sans', sans-serif"},
    marginTop: 45,
  })
```
